Dependently Typed Languages in Statix

### 2022-10-26 paper-evcs/paper.tex
Minor changes
Add \co formatting
Process Ola's feedback

% Commands used
\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }

@@

In this paper we show that Statix is already usable as a tool for implementing dependent types. There were some challenges in implementing the dependent type system, which this paper also discusses. Finally, we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.

@@

\section{Calculus of Constructions}

@@

We implement beta reduction using a Krivine abstract machine\cite{krivine}. This turned out to be the more natural way of expressing this over substitution-based evaluation relation. We originally tried to implement a substitution-based evaluation relation, which works fine for the base language. However, it runs into trouble when implementing inductive data types, more information about this will be in the full master thesis. Furthermore, abstract machines are usually more efficient than substitution-based approaches.

@@

We need to define multiple relations that will be used later for type-checking. First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.

Additionally, we define \verb|betaReduce| fully beta reduces a term. It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.

Finally, we define \verb|expectBetaEq|. This rule first beta reduces the heads of both sides, and then compares them. If the head is not the same, the rule fails. Otherwise, it recurses on the sub-expressions. One special case is when comparing two \verb|FnConstruct|s. Here we need to take into account alpha equality, that it, two expressions which only differ in the names that they use should be considered equal. We implement this by substituting in the body of the functions, replacing their argument names with placeholders.

\begin{figure}[h]
	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} { 
			\bhr
			{ \scope{s}{Type} }
			{ [] }
			{ \scope{s}{Type}} 
		}

		% Let rule
		\inferrule{ 
			\bhr 
			{ \scope{ sPutSubst(s, n, (s, v))}{b} } 
			{ t }
			{ \scope{ s' } { e' } }
		}{
			\bhr
			{ \scope{ s }{ Let(n, v, b) } }
			{ t }
			{ \scope { s' } { e' } }
		}

		% Var rule - NSubst
		\inferrule{ 
			sGetName(s, n) = NSubst(se, e) 
			\\ \bhr
			{\scope{se}{e}}
			{t}
			{\scope{se'}{e'}} 
		}{
			\bhr
			{\scope{s}{Var(n)}}
			{t}
			{\scope{se'}{e'}} 
		}

		% Var rule - NType
		\inferrule{ 
			sGetName(s, n) = NType(t) 
		}{ 
			\bhr
			{\scope{s}{Var(n)}}
			{t}
			{rebuild(s, Var(n), t)} 
		}

		% FnType rule
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{FnType(n, a, b)}}
			{[]}
			{\scope{s}{FnType(n, a, b)}} 
		}

		% FnConstruct rule - No args
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{FnConstruct(n, a, b)}}
			{[]}
			{\scope{s}{FnConstruct(n, a, b)}} 
		}

		% FnConstruct rule - Args
		\inferrule{ 
			\bhr
			{\scope{sPutSubst(s, n, a)}{b}}
			{ts}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{FnConstruct(n, a, b)}}
			{(t::ts)}
			{\scope{s'}{e'}} 
		}

		% FnDestruct rule
		\inferrule{ 
			\bhr
			{\scope{s}{f}}
			{(a::ts)}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{FnDestruct(f, a)}}
			{ts}
			{\scope{s'}{e'}} 
		}
	\end{mathpar}
	\caption{Rules for beta head reducing the Calculus of Constructions}
	\label{fig:beta-head-reduce-rules}
\end{figure}

@@

\begin{figure}[h]
	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} {
			\toe{s}{Type()}{Type()}
		}

		% Let rule
		\inferrule{ 
			\toe{s}{v}{vt}
			\\ \toe{sPutSubst(s, n, (s, v))}{b}{t}
		}{
			\toe{s}{Let(n, v, b)}{t}
		}

@@

		% Var rule - NType
		\inferrule{ 
			sGetName(s, n) = NType(t) 
		}{ 
			\toe{s}{Var(n)}{t}
		}

		% Var rule - NSubst
		\inferrule{ 
			sGetName(s, n) = NSubst(se, e) 
			\\ \toe{se}{e}{t}
		}{
		 	\toe{s}{Var(n)}{t}
		}

		% FnType rule
		\inferrule{ 
			\toe{s}{a}{at}
			\\ \beq{at}{Type()} 
			\\ \bred{s}{a}{a'}
			\\\\  \toe{sPutType(s, n, a')}{b}{bt}
			\\ \beq{bt}{Type()}
		} { 
			\toe{s}{FnType(n, a, b)}{Type()}
		}

@@

		% FnConstruct rule
		\inferrule{ 
			\toe{s}{a}{at}
			\\ \beq{at}{Type()}
			\\ \bred{s}{a}{a'} 
			\\\\ \toe{sPutType(s, n, a')}{b}{bt}
		} { 
			\toe{s}{FnConstruct(n, a, b)}{FnType(n, a', bt)}
		}

		% FnDestruct rule
		\inferrule{ 
			\toe{s}{f}{ft}
			\\ \bhr
				{\scope{s}{ft}}
				{[]}
				{\scope{s'}{FnType(da, dt, db)}} 
			\\\\ \toe{s}{a}{at} 
			\\ \beq{at}{\scope{sf}{dt}}
			\\ \bred{s'}{db}{db'}
		} { 
			\toe{s}{FnDestruct(f, a)}{db'}
	 	}

@@

\subsection{Avoiding variable capture}
One problem with the current implementation is that it does not avoid variable capture. Variable capture happens when a term becomes bound because of a wrong substitution. For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type. \T : T. T| is \verb|T : Type -> T : T -> T|. This is not the correct type, and even worse, the correct type is not possible to express without renaming! The problem is that there are multiple names, and there is no way to distinguish between them. This can be solved by using scopes to distinguish names.

\section{Extensions}
This section will discuss some further ideas that can be explored to build upon what has already been shown. These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.

@@

\paragraph*{Semantic Code Completion}
Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages. It functions works well, only showing completions that are semantically possible.

\section{Related Work}
The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope. In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is shown that does lazy substitutions, by using scopes as types. It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.

Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language. The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper. In principle, the implementations are very similar. For example, figure 3 of pi-forall is similar to figure \ref{fig:type-check-rules} from this paper. The primary difference is that they use a bidirectional type system, whereas this paper does not.

>>>>

% Commands used
\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }
\newcommand{\co}[1]{ \mathsf{#1} }

@@

In section \ref{sec:coc} of this paper we show that Statix is already usable as a tool for implementing dependent types. There were some challenges in implementing the dependent type system, which this paper also discusses. Finally, in section \ref{sec:extensions} we describe how to extend the language with several common features such as inductive data types, universes, and inference of implicit arguments.

@@

\section{Calculus of Constructions}
\label{sec:coc}

@@

We implemented beta reduction using a Krivine abstract machine\cite{krivine}. This turned out to be the more natural way of expressing this over substitution-based evaluation relation. We originally tried to implement the latter, which works fine for the base language. However, it runs into trouble when implementing inductive data types; more information about this will be in the full master thesis. An additional benefit is that abstract machines are usually more efficient than substitution-based approaches.

@@

We need to define multiple relations that will be used later for type-checking. First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from \cite{krivine}, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.

Additionally, we define \verb|betaReduce| which fully beta reduces a term. It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.

Finally, we define \verb|expectBetaEq|. This rule first beta reduces the heads of both sides, and then compares them. If the head is not the same, the rule fails. Otherwise, it recurses on the sub-expressions. One special case is when comparing two \verb|FnConstruct|s. Here we need to take into account alpha equality: two expressions which only differ in the names that they use should be considered equal. We implement this by substituting in the body of the functions, replacing their argument names with placeholders.

\begin{figure}[h]
	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} { 
			\bhr
			{ \scope{s}{\co{Type}()} }
			{ [] }
			{ \scope{s}{\co{Type}()}} 
		}

		% Let rule
		\inferrule{ 
			\bhr 
			{ \scope{ \co{sPutSubst}(s, n, (s, v))}{b} } 
			{ t }
			{ \scope{ s' } { e' } }
		}{
			\bhr
			{ \scope{ s }{ \co{Let}(n, v, b) } }
			{ t }
			{ \scope { s' } { e' } }
		}

		% Var rule - NSubst
		\inferrule{ 
			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
			\\ \bhr
			{\scope{se}{e}}
			{t}
			{\scope{se'}{e'}} 
		}{
			\bhr
			{\scope{s}{\co{Var}(n)}}
			{t}
			{\scope{se'}{e'}} 
		}

		% Var rule - NType
		\inferrule{ 
			\co{sGetName}(s, n) = \co{NType}(t) 
		}{ 
			\bhr
			{\scope{s}{\co{Var}(n)}}
			{t}
			{\co{rebuild}(s, \co{Var}(n), t)} 
		}

		% FnType rule
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{\co{FnType}(n, a, b)}}
			{[]}
			{\scope{s}{\co{FnType}(n, a, b)}} 
		}

		% FnConstruct rule - No args
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{\co{FnConstruct}(n, a, b)}}
			{[]}
			{\scope{s}{\co{FnConstruct}(n, a, b)}} 
		}

		% FnConstruct rule - Args
		\inferrule{ 
			\bhr
			{\scope{\co{sPutSubst}(s, n, a)}{b}}
			{ts}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{\co{FnConstruct}(n, a, b)}}
			{(t::ts)}
			{\scope{s'}{e'}} 
		}

		% FnDestruct rule
		\inferrule{ 
			\bhr
			{\scope{s}{f}}
			{(a::ts)}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{\co{FnDestruct}(f, a)}}
			{ts}
			{\scope{s'}{e'}} 
		}
	\end{mathpar}
	\caption{Rules for beta head reducing the Calculus of Constructions}
	\label{fig:beta-head-reduce-rules}
\end{figure}

@@

\begin{figure}[h]
	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} {
			\toe{s}{\co{Type}()}{\co{Type}()}
		}

		% Let rule
		\inferrule{ 
			\toe{s}{v}{vt}
			\\ \toe{\co{sPutSubst}(s, n, (s, v))}{b}{t}
		}{
			\toe{s}{\co{Let}(n, v, b)}{t}
		}

@@

		% Var rule - NType
		\inferrule{ 
			\co{sGetName}(s, n) = \co{NType}(t) 
		}{ 
			\toe{s}{\co{Var}(n)}{t}
		}

		% Var rule - NSubst
		\inferrule{ 
			\co{sGetName}(s, n) = \co{NSubst}(se, e) 
			\\ \toe{se}{e}{t}
		}{
		 	\toe{s}{\co{Var}(n)}{t}
		}

		% FnType rule
		\inferrule{ 
			\toe{s}{a}{at}
			\\ \beq{at}{\co{Type}()} 
			\\ \bred{s}{a}{a'}
			\\\\  \toe{\co{sPutType}(s, n, a')}{b}{bt}
			\\ \beq{bt}{\co{Type}()}
		} { 
			\toe{s}{\co{FnType}(n, a, b)}{\co{Type}()}
		}

@@

		% FnConstruct rule
		\inferrule{ 
			\toe{s}{a}{at}
			\\ \beq{at}{\co{Type}()}
			\\ \bred{s}{a}{a'} 
			\\\\ \toe{\co{sPutType}(s, n, a')}{b}{bt}
		} { 
			\toe{s}{\co{FnConstruct}(n, a, b)}{\co{FnType}(n, a', bt)}
		}

		% FnDestruct rule
		\inferrule{ 
			\toe{s}{f}{ft}
			\\ \bhr
				{\scope{s}{ft}}
				{[]}
				{\scope{s'}{\co{FnType}(da, dt, db)}} 
			\\\\ \toe{s}{a}{at} 
			\\ \beq{at}{\scope{sf}{dt}}
			\\ \bred{s'}{db}{db'}
		} { 
			\toe{s}{\co{FnDestruct}(f, a)}{db'}
	 	}

@@

\subsection{Avoiding variable capture}
One problem with the current implementation is that it does not avoid variable capture. Variable capture happens when a term becomes bound because of a wrong substitution. For example, according to the inference rules in figure \ref{fig:type-check-rules} the type of \verb|\T : Type. \T : T. T| is \verb|T : Type -> T : T -> T|. This is not the correct type, and even worse, the correct type is not possible to express without renaming! The problem is that there are multiple names, and there is no way to distinguish between them. This can be solved by using scopes to distinguish names. A full explanation of this will be in the master thesis.

\section{Extensions}
\label{sec:extensions}
This section will discuss some further ideas that can be explored to build upon what has already been shown. These will be fully described in the master thesis that this paper is based on, this is just a look into what is possible.

@@

\paragraph*{Semantic Code Completion}
Finally, we explored how semantic code completion presented in \cite{codecompletion} applies to dependently typed languages. It works well, only showing completions that are semantically possible.

For example, in the following code it would only suggest expressions that can be booleans:
\begin{lstlisting}
let f = \b: Bool. b;
f {{Expr}}
\end{lstlisting}

\section{Related Work}
The implementation in this paper requires performing substitutions in types immediately, as types don't have a scope. In section 2.5 of Scopes as Types \cite{scopes_as_types}, an implementation of System F is presented; it does lazy substitutions, by using scopes as types. It would be interesting to see if this approach could also apply to the Calculus of Constructions, where types can contain terms.

Another interesting comparison is to see how implementing a dependently typed language in Statix differs from implementing it in a general purpose language. The pi-forall language\cite{pi_forall} is a good example of a language with a similar complexity to the language presented in this paper. In principle, the implementations are very similar. For example, the inference rules presented in \cite{pi_forall} are similar to the inference rules presented in figure \ref{fig:type-check-rules} from this paper. The primary difference is that they use a bidirectional type system, whereas this paper does not.
